Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics

### 2016-10-15 Manuscripts_GMD.tex
update plume evolution section.

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementation of SPH. Core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash posed on aircrafts and communities. During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide sources terms for VATDs. This paper reports a new 3D (three dimensional) volcano plume model exploiting advantage of mesh-free methods in multiphase free boundary flow modeling.\\

@@

The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details through 1D numerical simulations. Recently developed 1D models (eg. \citep{mastin2007user, woodhouse2013interaction}) tend to account for more physics effects in one model. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational sources and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. Simutaneously, relative coarse particle might separate with main stream of the plume falling down onto the ground and may be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. Even entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation. 
All in all, the process of plume development is  essentially a multiphase turbulent mixing process coupled with heat transfer and many other microphysics and chemical reaction. As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and physics heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption). Because of above assumptions, all other micro-physics process (like phase change of  \texorpdfstring{H\textsubscript{2}O}, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process will not be considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refereed as weak plume. Our model is also lack ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano at April 4th \citep{sigurdsson19841982, folch2016fplume}. At present stage, we are more interested in more basic (and more critical) aspects of volcano plume and therefor devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods.\\

@@

As particles distribute in the space in a disordered manner. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The interpolation method is the basic SPH kernel based interpolation.

@@

@@

\includegraphics[width=5cm]{t15}
\caption{Contour of mass fraction and velocity quiver at 15 seconds, column rises up quickly due to initial momentum.}
\label{fig:strong_t15}

@@

@@

\caption{Contour of mass fraction and velocity quiver at 50 seconds, column keeps rising up and at the same time entraining air into the plume at the outer shear region. The entrainment of air is shown in a zoomed view in which the velocity quiver clearly shows entrainment of air into the plume at the margin of plume.}

@@

{
\includegraphics[width=10cm]{t100}
\caption{Contour of mass fraction and velocity quiver at 100 seconds, it keeps rising up and entraining air. The entrainment of air is clearly shown in zoomed view}
\label{fig:strong_plume_t100}
}
\includegraphics[width=5cm]{t200}
\caption{Contour of mass fraction at 200 seconds, plume reaches its top height}
\label{fig:strong_t200}
\includegraphics[width=5cm]{t300}
\caption{Contour of mass fraction at 300 seconds, the cloud is spreading}
\label{fig:strong_t300}
Figure \ref{fig:strong_t15}~\ref{fig:strong_t300} shows the dynamic evolution process of volcano plume. The contour level of mass fraction are 0.7, 0.5, 0.3, 0.1, $1\times 10 ^{-1}$, $1\times 10 ^{-2}$, $1\times 10 ^{-3}$, $1\times 10^{-4}$ and $1\times 10^{-5}$. The plume keeps rising up and entraining air into the plume until it reaches its top height(around t=200 seconds). Afterwards the mixtures of erupted material and air then fall back to neutral height and starts spreading radially. Detailed velocity quiver plots show entraining of air into the plume at the edge of column. The evolution of the plume further verifies our model.

@@

\authorcontribution{TEXT}

@@

All developers of Titan-2D, especially Dinesh Kumar who developed the GSPH version of Titan-2D, are greatly appreciated as Plume-SPH is based on their code. Advice and data (which is not shown in this paper) provided by Suzuki Yujiro gave us great help at the initial stage of model establishment and therefor are greatly appreciated. We appreciate Tomaso Esposti Ongaro for providing simulation data of PDAC (were also provided by Antonio Costa later together with other data) which helped us in doing early verification and improvement. We appreciate Antonio Costa for providing simulation results of existing 3D models, which are used in verification and validation section of this paper. We thank Matteo Cerminara for his helping on post-processing of plume simulation results. Computational results reported here were performed at the Center for Computational Research at the University at Buffalo. This project is supported by Grants No. NSF 1131074 from the National Science Foundation.

>>>>

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Effect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our concern in current work is the risks that volcanic ash poses on aircrafts and communities. During volcanic eruptions, VATDs (volcanic ash transport and dispersion models) are used to forecast the location and movement of ash clouds over hours to days. Those VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide source terms for VATDs. 
%This paper reports a new 3D (three dimensional) volcano plume model exploiting advantage of mesh-free methods in multiphase free boundary flow modeling.\\

@@

The effects of magma type and vent conditions \citep{woods1988fluid, woods1995decompression}, atmospheric conditions \citep{ woods1993moist, sparks1997volcanic, bursik2001effect}, external surface water \citep{koyaguchi1996formation}, thermal disequilibrium, and particle fallout \citep{woods1991particle} have been assessed in increasing details through 1D numerical simulations. Recently developed 1D models (eg. \citep{mastin2007user, woodhouse2013interaction}) tend to account for more physics effects in one model. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, In these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models requires little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcano plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skills in associated hazards forecasts with 1D models requires 2D (two dimensional) and 3D models.\\

@@

During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. When plume rises up, relative coarse particle might separate with main stream of the plume falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation. 
All in all, the process of plume development is  essentially a multiphase turbulent mixing process coupled with heat transfer and many other microphysics and chemical reaction.\\
As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and physics heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption). \\
Because of above assumptions, all other micro-physics process (like phase change of  \texorpdfstring{H\textsubscript{2}O}, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process is not be considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refereed as weak plume. Our model is also lack ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano at April 4th \citep{sigurdsson19841982, folch2016fplume}. At present stage, we are more interested in more basic (and more critical) aspects of volcano plume and therefor devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods.\\

@@

@@

As particles distribute in the space in a disordered manner. We first project simulation results (on disordered particles) onto a pre-defined grid before doing time average and spatial integration. The interpolation method is the basic SPH kernel based interpolation.\\

@@

\includegraphics[width=15cm]{strong_elevation}
\caption{This figure shows the dynamic evolution process of the Pinatubo with time. At the very beginning, it is driven by initial momentum and rises up quickly, mixing and entraining surround air. Around 100s, the jet stage finishes and it becomes a plume as there were enough air entrained. The plume keeps rising up until reaches the top height then starts spreading.}
\label{fig:strong_elevation}

@@

%\begin{figure}
%\center
%\includegraphics[width=5cm]{t15}
%\caption{Contour of mass fraction and velocity quiver at 15 seconds, column rises up quickly due to initial momentum.}
%\label{fig:strong_t15}
%\end{figure}

@@

\caption{Contour of mass fraction and velocity quiver at 50 seconds in plane x=0, the column keeps rising up and at the same time entraining air into the plume. The zoomed view (right figure) of the quiver plot clearly shows mechanism of air entrainment. The contour levels of mass fraction are 0.7, 0.5, 0.3, 0.1, $1\times 10 ^{-1}$, $1\times 10 ^{-2}$, $1\times 10 ^{-3}$, $1\times 10^{-4}$ and $1\times 10^{-5}$.}

@@

%{
%\includegraphics[width=10cm]{t100}
%\caption{Contour of mass fraction and velocity quiver at 100 seconds. The entrainment of air is clearly shown in the zoomed view (right figure). }
%\label{fig:strong_plume_t100}
%}
%\includegraphics[width=5cm]{t200}
%\caption{Contour of mass fraction at 200 seconds, plume reaches its top height}
%\label{fig:strong_t200}
%\includegraphics[width=5cm]{t300}
%\caption{Contour of mass fraction at 300 seconds, the cloud is spreading}
%\label{fig:strong_t300}
Figure \ref{fig:strong_elevation} shows the overall evolution process of plume. Plume keeps rising up and entraining air into the plume until it reaches its top height(around t=300 seconds). Afterwards the mixtures of erupted material and air then fall back to neutral height and starts spreading radially. Detailed velocity quiver plot in Fig. \ref{fig:strong_plume_t50} shows entraining of air into the plume at the edge of column. The evolution of the plume further verifies our model.

@@

%\authorcontribution{TEXT}

@@

All developers of Titan-2D, especially Dinesh Kumar who developed the GSPH version of Titan-2D, are greatly appreciated as Plume-SPH is based on their code. Advice and data (which is not shown in this paper) given by Suzuki Yujiro gave us great help at the initial stage of model establishment and therefor are greatly appreciated. We appreciate Tomaso Esposti Ongaro for providing simulation data of PDAC (were also provided by Antonio Costa later together with simulation results of other plume models) which helped us in doing early verification and improvement. We appreciate Antonio Costa for providing simulation results of existing 3D models, which are used in verification and validation section of this paper. We thank Matteo Cerminara for his helping on post-processing of plume simulation results. Computational results reported here were performed at the Center for Computational Research at the University at Buffalo. This project is supported by Grants No. NSF 1131074 from the National Science Foundation.
